---
title: "Reproducible Research: Peer Assessment 2"
author: "David Bradford"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 html_document:
   keep_md: true
   #code_folding: hide
   #toc: true
   #toc_float: true
   #theme: cosmo
---

## Synopsis



```{r load packages}
library(ggplot2)
library(plyr)
library(dplyr)
library(data.table)
library(grid)
```

## Data Processing

First, read in the US NOAA storm database located here:  https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Then unzip the .csv into the working /data directory and read into the storms data frame.

```{r load data}
# set working & create data directory
setwd("C:\\Users\\dabradford\\Desktop\\Coursera\\DataSci\\Reproducible\\RepData_PeerAssessment2")
if(!file.exists("./data")){dir.create("./data")}

# download storm data
get.data.project <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(get.data.project,destfile="./data/storm_data.csv.bz2",method="auto")

zipfile.data = "storm_data.csv.bz2"

# make sure the data is in the working directory if not download the zip file into the to zipfile.data and unzip it
if(!file.exists(zipfile.data)) {        
    unzip(zipfile="./data/storm_data.csv.bz2",exdir="./data")
} 

path_act <- file.path("./data" , "./data/storm_data")
files<-list.files(path_act, recursive=TRUE)

# Read data file > THIS WILL TAKE A WHILE
#storms_raw <- read.csv("./data/storm_data.csv.bz2")

# inspect data file
str(storms_raw)
summary(storms_raw)
```

This analysis specifically asks for the effects of storm events across the United States.  For this analysis we will not be considering geographical boundaries, such as county or state borders, and will include all geolocations included in the dataset.  Additionally, we will only be looking at the last ten years of data for our analysis to account for more recent trends and to ensure we are working with more accurate data points.  

Filter data to only include events from November 2001 through November 2011 based on the __BGN_DATE__ variable.

```{r}
# First convert BGN_DATE to date 
storms_raw$BGN_DATE <- as.Date(storms_raw$BGN_DATE, "%m/%d/%Y")

# Then parse only those events between Nov 2001 and Nov 2011 to get the last ten years of data.  Also only include the variables that will help us understand consequences to population health (FATALITIES & INJURIES) and/or economics (PROPDMG &  CROPDMG)

storms_10yr  <- storms_raw %>%
    
    filter(BGN_DATE >= "2001-11-01" & BGN_DATE <= "2011-11-30")
```

We need to understand the dataset better since there isn't clear documentation of what some of the variable contents signify.  Let's view a few events in Georgia

```{r}
# view one event to determine variable meanings
ga_events  <- storms_raw %>%
    select(BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
    filter(BGN_DATE == "2007-03-01" & EVTYPE == "TORNADO" & STATE == "GA")

head(ga_events)
```

Based on this analysis and comparing to the NOAA Storm Events database search page at the link below we understand the following:

__FATALITIES__ - Number of people killed as a direct result of the event
__INJURIES__ - Number of people injured as a direct result of the event

__Property damage__ is found by combining PROPDMG and PROPDMGEXP variables.  Possible values for PROPDMGEXP include:

<blank> - no damage
0 - no damage
B - damage in billions USD.  For example, if PROPDMG = 5, and PROPDMG = 'B' then property damage is $5,000,000,000
K - damage in thousands USD
M - damage in millions USD

__Crop damage__ is calculated in the same way.

https://www.ncdc.noaa.gov/stormevents/choosedates.jsp?statefips=-999%2CALL

```{r}
count(storms_10yr, 'PROPDMGEXP')
count(storms_10yr, 'CROPDMGEXP')
```
 
Therefore we need to add to new variables to account for property and crop damage.

```{r}
# Property damage
storms_10yr <- mutate(storms_10yr,
                      prop.dmg = ifelse(PROPDMGEXP == "K", PROPDMG * 1000,
                                        ifelse(PROPDMGEXP == "M", PROPDMG * 1000000,
                                               ifelse(PROPDMGEXP == "B", PROPDMG * 1000000000, 0))))
# Crop damage
storms_10yr <- mutate(storms_10yr,
                      crop.dmg = ifelse(CROPDMGEXP == "K", CROPDMG * 1000,
                                        ifelse(CROPDMGEXP == "M", CROPDMG * 1000000,
                                               ifelse(CROPDMGEXP == "B", CROPDMG * 1000000000, 0))))
```

Since there are 128 event types we wanted to plot the top 20 and categorize all other events as "OTHER".

```{r}
top10fatalities <- data.table(storms_10yr)
top10fatalities <- top10fatalities[FATALITIES != 0 & !is.na(FATALITIES), sum(FATALITIES), by = EVTYPE]

top10fatalities <- top10fatalities[order(-rank(V1), EVTYPE)[21:nrow(top10fatalities)],EVTYPE:="OTHER*"]
top10fatalities <- top10fatalities[, sum(V1), by = EVTYPE]
top10fatalities$EVTYPE <- factor(top10fatalities$EVTYPE, levels = top10fatalities$EVTYPE[order(top10fatalities$V1)])


top10injuries <- data.table(storms_10yr)
top10injuries <- top10injuries[INJURIES != 0 & !is.na(INJURIES), sum(INJURIES), by = EVTYPE]

top10injuries <- top10injuries[order(-rank(V1), EVTYPE)[21:nrow(top10injuries)],EVTYPE:="OTHER*"]
top10injuries <- top10injuries[, sum(V1), by = EVTYPE]
top10injuries$EVTYPE <- factor(top10injuries$EVTYPE, levels = top10injuries$EVTYPE[order(top10injuries$V1)])
```



## Results

### Q1 - Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r}

# create png device
png("plot1.png", width=960, height=480, units="px", bg="transparent")

plot1 <- ggplot(top10fatalities, aes(x=EVTYPE, y=V1)) + 
    geom_bar(stat="identity", fill="blue") +
    labs(x="Event Type", y="# Fatalities", title="# Fatalities by Event Type (Top 20)") +
    theme_minimal() +
    coord_flip() + 
    geom_text(aes(label=V1), size=4, position=position_dodge(width=1), hjust=-.5) +
    scale_y_continuous(breaks = seq(min(0), max(1500), by = 200)) 

# turn off device to ensure writing completes
dev.off()

plot1
```

```{r}
# create png device
png("plot2.png", width=960, height=480, units="px", bg="transparent")

plot2 <- ggplot(top10injuries, aes(x=EVTYPE, y=V1)) + 
    geom_bar(stat="identity", fill="blue") +
    labs(x="Event Type", y="# Injuries", title="# Injuries by Event Type (Top 20)") +
    theme_minimal() +
    coord_flip() + 
    geom_text(aes(label=V1), size=4, position=position_dodge(width=1), hjust=-.5) +
    scale_y_continuous(breaks = seq(min(0), max(1500), by = 200)) 

# turn off device to ensure writing completes
dev.off()

plot2
```



### Q2 - Across the United States, which types of events have the greatest economic consequences?


## Appendix



