---
title: "Reproducible Research: Peer Assessment 2"
author: "David Bradford"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
 html_document:
   keep_md: true
   #code_folding: hide
   #toc: true
   #toc_float: true
   #theme: cosmo
---

## Synopsis



```{r load packages, echo=TRUE, message=FALSE, warning=FALSE}
#library(ggplot2)
#library(dplyr)
#library(plyr)
#library(data.table)
#library(grid)
#library(gridExtra)
#library(lubridate)
```

## Data Processing

First, read in the US NOAA storm database located here:  https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Then unzip the .csv into the working /data directory and read into the storms data frame.

```{r load data message=FALSE, warning=FALSE, echo=TRUE}
# set working & create data directory
setwd("C:\\Users\\dabradford\\Desktop\\Coursera\\DataSci\\Reproducible\\RepData_PeerAssessment2")
if(!file.exists("./data")){dir.create("./data")}

# download storm data
get.data.project <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(get.data.project,destfile="./data/storm_data.csv.bz2",method="auto")

zipfile.data = "storm_data.csv.bz2"

# make sure the data is in the working directory if not download the zip file into the to zipfile.data and unzip it
if(!file.exists(zipfile.data)) {        
    unzip(zipfile="./data/storm_data.csv.bz2",exdir="./data")
} 

path_act <- file.path("./data" , "./data/storm_data")
files<-list.files(path_act, recursive=TRUE)

# Read data file
storms <- read.csv("./data/storm_data.csv.bz2")


# inspect data file
str(storms)
summary(storms)
```

## Results

### Q1 - Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?



### Q2 - Across the United States, which types of events have the greatest economic consequences?


## Appendix



